\bf Model Visualization Techniques for a Social Network Model

### 2018-07-25 paper-child.Rnw
getting there...

@@

The CTMC process $X(T)$ is a series of single tie changes that happen according to a pre-defined rate function, where one actor at a time is selected and it adds or removes one outgoing tie, or does nothing. Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize its utility function based on the current and possible future states of the network.

\textbf{The Rate Function} In general, each actor $i$ in the network $x$ may change its ties, $x_{ij}$, to other nodes $j \neq i$ according to the SAOM's  \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$. With this general formulation, it is possible that each node will have a different rate of change. We assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between time $t_m$ and $t_{m+1}$. With this simple rate function, we have just one set of rate parameters in the overall model. In this simple model, the rate parameters dictate how often an actor $i$ ``gets an opportunity" to change one of its ties, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$ (Note that when $j = i$, no change is made.) We also assume that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state. Let $\tau(i|x,m) $ be the wait time until actor $i$ gets to the opportunity to change its current set of ties. For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change made by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$. Then, we express this in Equation~\ref{eq:c1}.

@@

From Equation~\ref{eq:c1}, we can derive the waiting time to the next change opportunity by \textit{any} actor in the network, $\tau(x)$. The value $\tau(x)$ is also exponentially distributed with expected value $(n\alpha_m)^{-1}$, where $n\alpha_m$ is the rate at which any tie change occurs. The estimation of this parameter in \texttt{RSiena} is simple: the method of moments is used to estimate the rate with the statistic 
\begin{equation}\label{eq:cdiff}
C = \sum_i\sum_j |x_{ij}(t_{m+1}) - x_{ij}(t_m)|
\end{equation}
which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$.

\textbf{The Objective Function} Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is changing at a time in the CTMC. The node $i$, which is the node that is given the opportunity to change at the current time point, is called the \textit{ego} node. It has the potential to interact with all other nodes in the network, $j \neq i$. These nodes $j$, are referred to as \textit{alter} nodes, and do not change any of their ties when $i$ may change. For the ego node $i$ in the current network state $x$, its \emph{objective function}, which it tries to maximize, is written as
\begin{equation} \label{eq:of1}
f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}
for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$, the matrix of covariates. The vector $\boldsymbol{\beta}$ contains the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$. To increase the value of $f_i$, $i$ will destroy one tie where $x_{ij} = 1$, create one tie where $x_{ij}=0$, or make no change if any changes result in lower values of $f_i$.

@@

<<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data. The subset we will be using is outlined in red.", results='hide',fig.show='asis', fig.width=8, fig.height=3, eval = FALSE>>=
source("code/00e_small_friends.R")
view_rs2$wave <- factor(view_rs2$wave)
levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
ggplot(data= view_rs2, aes(x = y, y=-x)) +
  geom_tile(aes(fill = as.factor(value))) +
  scale_fill_manual("i likes j", labels=c("no", "yes"),
                    values = c('white', 'black')) +
  geom_rect(data= NULL, inherit.aes = FALSE, color = 'red',
            aes(xmin = 19.5, xmax = 35.5, ymax = -19.5, ymin = -35.5, fill =NA)) +
  facet_wrap(~wave) + theme_bw() +
  labs(x = "node j (alter)", y = "node i (ego)") +
  ThemeNoNet +
  theme(aspect.ratio=1, axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())
@

<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper. The nodes change color as their drinking behavior changes, and the network becomes less densely connected over time. These are the mechanisms we hope to capture with SAOMs.", out.width='.99\\textwidth', fig.height=3>>=
source("code/00e_small_friends.R")
view_rs2$wave <- factor(view_rs2$wave)
levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))),
                 data.frame(id = 1:16, drink = drink2[,1]),
                 by.x = "X1", by.y = "id", all = T)
actual1$Wave <- 1
actual2 <- merge(data.frame(as.edgelist(as.network(fd2.w2))),
                 data.frame(id = 1:16,drink = drink2[,2]),
                 by.x = "X1", by.y = "id", all = T)
actual2$Wave <- 2
actual3 <- merge(data.frame(as.edgelist(as.network(fd2.w3))),
                 data.frame(id = 1:16,drink = drink2[,3]),
                 by.x = "X1", by.y = "id", all = T)
actual3$Wave <- 3
alldat <- rbind(actual1, actual2, actual3)
#alldat$X1 <- paste0("V", alldat$X1)
alldat$X1 <- as.factor(alldat$X1)
#alldat$X2 <- ifelse(!is.na(alldat$X2), paste0("V", alldat$X2), NA)
alldat$X2 <- as.factor(alldat$X2)
#write_csv(alldat, "smallfriends4Geomnet.csv")
alldataviz <- ggplot(data = alldat,
       aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 5, arrowsize = .25, arrowgap = .05, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 4, vjust = .5, linewidth=0.5) +
  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) +
  ThemeNet +
  facet_wrap(~Wave, labeller = "label_both") +
  theme(legend.position = 'bottom',
        panel.background = element_rect(fill = NA, color = "grey30"),
        legend.text = element_text(size = 12, family = "Times New Roman"),
        legend.title = element_text(size = 12, family = "Times New Roman"))
alldataviz
@

To guide our visual exploration of SAOMs, we use two example data sources. The first is a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study" that is provided on the \texttt{RSiena} webpage. These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make node-link diagrams easier to see and to make any changes in the network more noticeable. %To determine which subset to select, we visualized all waves of the full network using the adjacency matrix visualization approach, which we show in Figure~\ref{fig:adjmatviz}. This adjacency matrix visualization is different from the one in Figure~\ref{fig:toyplots} because it does not show the node IDs on the axes. The network in Figure~\ref{fig:adjmatviz} is much larger than our toy data example, so we remove the node labels to remove clutter. In all three adjacency matrices, the ego and the alter nodes are ordered by node ID, 1-50, which were determined arbitrarily by the relationships in wave 1. 
%In Figure~\ref{fig:adjmatviz}, we show the adjacency matrix visualization of all three observations of this network, with the subset we selected for analysis outlined in red. This subset contains 16 actors, their ties and their drinking behavior for three timepoints. 
The node-link visualization of this data is provided in Figure~\ref{fig:alldataviz}. For model fitting, we condition on wave 1 and estimate the parameters of the models for the second and third waves. The actor-level categorical covariate, drinking behavior, has four values: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, and (4) drinks once a week. In Figure~\ref{fig:alldataviz}, the nodes are colored according to the drinking behavior of that student. Over time, we can see that the students tend to drink more and become increasingly isolated into smaller groups. An analysis of this type of data with a SAOM should capture these dynamics in a way that allows the researcher to draw conclusions about the nature of the network and behavioral forces at play.

@@

\subsection{Models}

We first consider the 16 actor subset of the teenage friends and lifestyle data show in Figure~\ref{fig:alldataviz}. To this data, we fit three different SAOMs. Each SAOM used a simple rate function, $\alpha_m$, and an objective function with two or three parameters. The first model, M1, contains the absolute minimum number of parameters in the objective function $f_i(x)$: 
\begin{equation}\label{eq:m1}
f_i(x)^{M1} = \beta_1s_{i1} + \beta_2s_{i2},
\end{equation}
where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$. The second and third models, M2 and M3, contain one additional parameter each in the objective function which were determined by a Wald-type test provided in the \texttt{RSiena} software to be significant, with $p$-values less than 0.05 \citep{RSienaTest}. The M2 model contains an actor-level covariate parameter, and the M3 model contains an additional strutural effect in the objective function.
\begin{align}\label{eq:m2m3}
f_i(x)^{M2} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_3s_{i3} \\
f_i(x)^{M3} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_4s_{i4},
\end{align}

@@

In Section~\ref{sec:samemoddiffdat}, we also fit models M1, M2, and M3 to the senate collaboration data for comparison. To fit M2 to the senate data, we use the number of bills authored by each senator as the node covariate for the jumping transitive triplet variable. %In terms of the senate data, then, the value of $\beta_3$ should dictate how willing a senator is to coauthor a bill on which the author of the bill has a different level of authorship, assuming there is an intermediary between the two who has the same authorship level as the first. %Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}.
%In addition, we fit two more models, M4 and M5 to the senate data that were found to be significant according to the Wald-type test:

@@

clintonSenate %>% filter(!(source %in% c("Roland Burris", "Bernard Sanders", "Frank R. Lautenberg", "Mary L. Landrieu") & is.na(target))) %>%
ggplot() +
  geom_net(directed = T, labelon=F, arrowsize = .3, singletons= F, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au)) +
  scale_size_continuous(name="# bills\nauthored") +
  #theme_net() +
  ThemeNet +
  scale_shape_manual(values = c(17,16)) +
  scale_color_manual(values = c("royalblue", "forestgreen","firebrick")) +
  theme(legend.position = 'bottom') +
  facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')
@

@@

<<summnetM1, message=FALSE, fig.cap='On the left, the first wave of observed data that is conditioned on in the model. On the right, the second wave of observed data. In the middle, a summary network from the first model fit to the data. This summary network represents 1,000 simulations of wave 2 using the values from the simple fitted model M1.',results='hide', fig.height=2.5>>=
friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
drink1 <- drink[20:35,1]
drink2 <- drink[20:35,2]
drink3 <- drink[20:35,3]
friendship <- sienaDependent(friendshipData)
alcohol <- varCovar(cbind(drink1,drink2, drink3))
mydata <- sienaDataCreate(friendship, alcohol)
M1eff <- getEffects(mydata)
M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
M3eff <- includeEffects(M1eff, nbrDist2twice)
simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
means <- simu2 %>% group_by(Model, parameter) %>% summarize(
  means = mean(estimate)
)
M1parms <- (means %>% filter(Model == "M1"))$means
M2parms <- (means %>% filter(Model == "M2"))$means
M3parms <- (means %>% filter(Model == "M3"))$means
set.seed(4231352)
#M1sims1000 <- netvizinf::saom_simulate(dat = mydata,
#                            struct = M1eff,
#                       parms = as.numeric(M1parms),
#                       N = 1000)
#save(M1sims1000, file = "data/M1sims1000.RData")
load("data/M1sims1000.RData")
# M2sims1000 <- netvizinf::saom_simulate(dat = mydata,
#                             struct = M2eff,
#                        parms = as.numeric(M2parms),
#                        N = 1000)
# save(M2sims1000, file = "data/M2sims1000.RData")
load("data/M1sims1000.RData")
# M3sims1000 <- netvizinf::saom_simulate(dat = mydata,
#                             struct = M3eff,
#                        parms = as.numeric(M3parms),
#                        N = 1000)
# save(M3sims1000, file = "data/M3sims1000.RData")
load("data/M3sims1000.RData")

@@

diff1 <- am_t1 - fd2.w1
diff1[which(diff1==1)] <- 2
diff1[which(fd2.w1 == 1 & am_t1 == 1)]  <- 1
diff1 <- diff1 %>% dplyr::as_data_frame() 
names(diff1) <- paste0("V", 1:16)
diff1 %>% mutate(from =paste0("V", 1:16)) %>% gather(to, value, -from) %>%
  mutate(from = as.character(parse_number(from)), to = as.character(parse_number(to)), 
         var = recode(value, `0` = "Neither", `2` = "InAvgNotW1",  `1`= "InBoth"),
         color2 = recode(value, `0`="white", `1`= "black", `2` = "green")) %>%
  filter(value !=0) %>% 
  full_join(t1, c("from","to")) -> t1_1 
t1_1 <- arrange(t1_1, from, to)
p1 <- ggplot(data = t1_1) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = t1_1$color2,
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
           singletons = T) +
  labs(title = "Difference between Wave 1 and Average from M1") + ThemeNet +
  theme(plot.background = element_rect(color = 'black'))

diff2 <- -am_t1 + fd2.w2 
diff2[which(diff2==1)] <- 2
diff2[which(fd2.w2 == 1 & am_t1 == 1)]  <- 1
diff2 <- dplyr::as_data_frame(diff2)
names(diff2) <- paste0("V", 1:16)
diff2 %>% mutate(from =paste0("V", 1:16)) %>% gather(to, value, -from) %>% 
  mutate(from = as.character(parse_number(from)), to = as.character(parse_number(to)), 
         var = recode(value, `0` = "Neither", `-1` = "InAvgNotW2",  `1`= "InBoth", `2` = "InW2notAvg"),
         color2 = recode(value, `0`="white", `1`= "black", `2` = "red", `-1` = "green")) %>% 
  filter(value !=0) %>% 
  full_join(t1, c("from","to")) -> t1_2 
t1_2$linewidth[is.na(t1_2$linewidth) & !is.na(t1_2$to)] <- 1
t1_2 <- arrange(t1_2, from, to)
t1_2 <- t1_2 %>% filter(!(from == 16 & is.na(to)))
p2 <- ggplot(data = t1_2) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = t1_2$color2,
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
           singletons = T) +
  labs(title = "Difference between Wave 2 and Average from M1") + ThemeNet +
  theme(plot.background = element_rect(color = 'black'))
grid.arrange(p1, p2, nrow = 1)
@

@@

<<gof2, fig.cap="An example of viewing the distribution of a statistic of interest from a SAOM. Here, the Mahalanobis distance of the cumulative outdegree values computed on a simulated network to the mean value simulated is shown. The distance computed from the data is shown as a vertical line.", fig.height=3, out.width='.6\\textwidth', eval = FALSE>>=
load("data/ansnullpaper.rda")
library(lattice)
gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
histdat <- data.frame(x = log(gof1$`Period 1`$SimulatedTestStat))
ggplot(histdat) + geom_histogram(aes(x = x), binwidth = .1, fill = 'white', color = "black") +
  #geom_density(aes(x = x, y = ..density..*250)) +
  geom_vline(xintercept = log(gof1$`Period 1`$ObservedTestStat), color = 'red') +
  ThemeNoNet +
  labs(x = "log(Statistic Value)", title = "Distribution of log Mahalanobis Distance")

@@

\section{Collections of models}

>>>>

<<getdata , echo=FALSE>>=
set.seed(823746)
load("data/ansnullpaper.rda")
ansnullchains <- get_chain_info(ansnull)
load("data/M1sims1000.RData")
load("data/M2sims1000.RData")
load("data/M3sims1000.RData")
@

@@

The CTMC process $X(T)$ is a series of single tie changes that happen according to a pre-defined rate function, where one actor at a time is selected and it adds or removes one outgoing tie, or does nothing. Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize its utility function based on the current and possible future states of the network. \\

\noindent \textbf{The Rate Function} In general, each actor $i$ in the network $x$ may change its ties, $x_{ij}$, to other nodes $j \neq i$ according to the SAOM's  \emph{rate function}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$. With this general formulation, it is possible that each node will have a different rate of change. We assume a simple rate function, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between time $t_m$ and $t_{m+1}$. With this simple rate function, we have just one set of rate parameters in the overall model. In this simple model, the rate parameters dictate how often an actor $i$ ``gets an opportunity" to change one of its ties, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$ for $m = 1, \dots, M-1$ (Note that when $j = i$, no change is made.) We also assume that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state. Let $\tau(i|x,m) $ be the wait time until actor $i$ gets to the opportunity to change its current set of ties. For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change made by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$. Then, we express this in Equation~\ref{eq:c1}.

@@

From Equation~\ref{eq:c1}, we can derive the waiting time to the next change opportunity by \textit{any} actor in the network, $\tau(x)$. The value $\tau(x)$ is also exponentially distributed with expected value $(n\alpha_m)^{-1}$, where $n\alpha_m$ is the rate at which any tie change occurs. The estimation of this parameter in \texttt{RSiena} is simple: the method of moments is used to estimate the rate with the statistic 
\begin{equation}\label{eq:cdiff}
C = \sum_i\sum_j |x_{ij}(t_{m+1}) - x_{ij}(t_m)|
\end{equation}
which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$. \\

\noindent \textbf{The Objective Function} Because of the conditional independence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is changing at a time in the CTMC. The node $i$, which is the node that is given the opportunity to change at the current time point, is called the \textit{ego} node. It has the potential to interact with all other nodes in the network, $j \neq i$. These nodes $j$, are referred to as \textit{alter} nodes, and do not change any of their ties when $i$ may change. For the ego node $i$ in the current network state $x$, its \emph{objective function}, which it tries to maximize, is written as
\begin{equation} \label{eq:of1}
f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}
for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$, the matrix of covariates. The vector $\boldsymbol{\beta}$ contains the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$. To increase the value of $f_i$, $i$ will destroy one tie where $x_{ij} = 1$, create one tie where $x_{ij}=0$, or make no change if any changes result in lower values of $f_i$.

@@

% removed the following plot
<<adjmatviz, fig.align='center', fig.cap="A visualization of the adjacency matrices of the three waves of network observations in the ``Teenage Friends and Lifestyle Study'' data. The subset we will be using is outlined in red.", results='hide',fig.show='asis', fig.width=8, fig.height=3, eval = FALSE>>=
source("code/00e_small_friends.R")
view_rs2$wave <- factor(view_rs2$wave)
levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
ggplot(data= view_rs2, aes(x = y, y=-x)) +
  geom_tile(aes(fill = as.factor(value))) +
  scale_fill_manual("i likes j", labels=c("no", "yes"),
                    values = c('white', 'black')) +
  geom_rect(data= NULL, inherit.aes = FALSE, color = 'red',
            aes(xmin = 19.5, xmax = 35.5, ymax = -19.5, ymin = -35.5, fill =NA)) +
  facet_wrap(~wave) + theme_bw() +
  labs(x = "node j (alter)", y = "node i (ego)") +
  ThemeNoNet +
  theme(aspect.ratio=1, axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())
@

<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper. The nodes change color as their drinking behavior changes, and the network becomes less densely connected over time. These are the mechanisms we hope to capture with SAOMs.", out.width='.99\\textwidth', fig.height=3, results='hide'>>=
source("code/00e_small_friends.R")
view_rs2$wave <- factor(view_rs2$wave)
levels(view_rs2$wave) <- c("Wave 1", "Wave 2", "Wave 3")
actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))),
                 data.frame(id = 1:16, drink = drink2[,1]),
                 by.x = "X1", by.y = "id", all = T)
actual1$Wave <- 1
actual2 <- merge(data.frame(as.edgelist(as.network(fd2.w2))),
                 data.frame(id = 1:16,drink = drink2[,2]),
                 by.x = "X1", by.y = "id", all = T)
actual2$Wave <- 2
actual3 <- merge(data.frame(as.edgelist(as.network(fd2.w3))),
                 data.frame(id = 1:16,drink = drink2[,3]),
                 by.x = "X1", by.y = "id", all = T)
actual3$Wave <- 3
alldat <- rbind(actual1, actual2, actual3)
#alldat$X1 <- paste0("V", alldat$X1)
alldat$X1 <- as.factor(alldat$X1)
#alldat$X2 <- ifelse(!is.na(alldat$X2), paste0("V", alldat$X2), NA)
alldat$X2 <- as.factor(alldat$X2)
#write_csv(alldat, "smallfriends4Geomnet.csv")
alldataviz <- ggplot(data = alldat,
       aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 5, arrowsize = .25, arrowgap = .05, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 4, vjust = .5, linewidth=0.5) +
  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) +
  ThemeNet +
  facet_wrap(~Wave, labeller = "label_both") +
  theme(legend.position = 'bottom',
        panel.background = element_rect(fill = NA, color = "grey30"),
        legend.text = element_text(size = 12, family = "Times New Roman"),
        legend.title = element_text(size = 12, family = "Times New Roman"))
alldataviz
@

\noindent \textbf{Example Data:} To guide our visual exploration of SAOMs, we use two example data sources. The first is a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study" that is provided on the \texttt{RSiena} webpage. These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make node-link diagrams easier to see and to make any changes in the network more noticeable. %To determine which subset to select, we visualized all waves of the full network using the adjacency matrix visualization approach, which we show in Figure~\ref{fig:adjmatviz}. This adjacency matrix visualization is different from the one in Figure~\ref{fig:toyplots} because it does not show the node IDs on the axes. The network in Figure~\ref{fig:adjmatviz} is much larger than our toy data example, so we remove the node labels to remove clutter. In all three adjacency matrices, the ego and the alter nodes are ordered by node ID, 1-50, which were determined arbitrarily by the relationships in wave 1. 
%In Figure~\ref{fig:adjmatviz}, we show the adjacency matrix visualization of all three observations of this network, with the subset we selected for analysis outlined in red. This subset contains 16 actors, their ties and their drinking behavior for three timepoints. 
The node-link visualization of this data is provided in Figure~\ref{fig:alldataviz}. For model fitting, we condition on wave 1 and estimate the parameters of the models for the second and third waves. The actor-level categorical covariate, drinking behavior, has four values: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, and (4) drinks once a week. In Figure~\ref{fig:alldataviz}, the nodes are colored according to the drinking behavior of that student. Over time, we can see that the students tend to drink more and become increasingly isolated into smaller groups. An analysis of this type of data with a SAOM should capture these dynamics in a way that allows the researcher to draw conclusions about the nature of the network and behavioral forces at play.

@@

\noindent \textbf{Models:} We first consider the 16 actor subset of the teenage friends and lifestyle data show in Figure~\ref{fig:alldataviz}. To this data, we fit three different SAOMs. Each SAOM used a simple rate function, $\alpha_m$, and an objective function with two or three parameters. The first model, M1, contains the absolute minimum number of parameters in the objective function $f_i(x)$: 
\begin{equation}\label{eq:m1}
f_i(x)^{M1} = \beta_1s_{i1} + \beta_2s_{i2},
\end{equation}
where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$. The second and third models, M2 and M3, contain one additional parameter each in the objective function which were determined by a Wald-type test provided in the \texttt{RSiena} software to be significant, with $p$-values less than 0.05 \citep{RSienaTest}. The M2 model contains an actor-level covariate parameter, and the M3 model contains an additional strutural effect in the objective function.
\begin{align}\label{eq:m2m3}
f_i(x)^{M2} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_3s_{i3} \\
f_i(x)^{M3} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_4s_{i4},
\end{align}

@@

In this paper, we use three different model visualization techniques on SAOMs. First, Section~\ref{sec:minds} places the models in the data space. Then, Section~\ref{sec:collectmodels} explores visualizations of various collections of models. Finally, in Section~\ref{sec:algorithms} we use model visualization to explore the underlying algorithms in the SAOM fitting process. 
 %In terms of the senate data, then, the value of $\beta_3$ should dictate how willing a senator is to coauthor a bill on which the author of the bill has a different level of authorship, assuming there is an intermediary between the two who has the same authorship level as the first. %Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}.
%In addition, we fit two more models, M4 and M5 to the senate data that were found to be significant according to the Wald-type test:

@@

clintonSenate %>% filter(!(source %in% c("Roland Burris", "Bernard Sanders", "Frank R. Lautenberg", "Mary L. Landrieu") & is.na(target))) %>%
ggplot() +
  geom_net(directed = T, labelon=F, arrowsize = .3, singletons= F, fiteach = T, arrowgap = .01, layout.alg = 'fruchtermanreingold',
           aes(from_id = source, to_id = target, color = party, linewidth = gsw, shape = sex,size = n_au, group = party)) +
  scale_size_continuous(name="# bills\nauthored" , guide = guide_legend(nrow= 2)) +
  #theme_net() +
  ThemeNet +
  scale_shape_manual(values = c(17,16), guide = guide_legend(nrow= 2)) +
  scale_color_manual(values = c("royalblue", "forestgreen","firebrick"), guide = guide_legend(nrow=2)) +
  facet_wrap(~Clinton, nrow = 1, labeller = 'label_both')  +
  theme(legend.position = "bottom")
@

@@

<<summnetM1, message=FALSE, fig.cap='On the left, the first wave of observed data that is conditioned on in the model. On the right, the second wave of observed data. In the middle, a summary network from the first model fit to the data. This summary network represents 1,000 simulations of wave 2 using the values from the simple fitted model M1.',results='hide', fig.height=2.5>>=
friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
friendshipData <- array(c(fd2.w1, fd2.w2, fd2.w3), dim = c(16, 16, 3))
drink1 <- drink[20:35,1]
drink2 <- drink[20:35,2]
drink3 <- drink[20:35,3]
friendship <- sienaDependent(friendshipData)
alcohol <- varCovar(cbind(drink1,drink2, drink3))
mydata <- sienaDataCreate(friendship, alcohol)
M1eff <- getEffects(mydata)
M2eff <- includeEffects(M1eff, jumpXTransTrip, interaction1 = "alcohol")
M3eff <- includeEffects(M1eff, nbrDist2twice)
simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")
means <- simu2 %>% group_by(Model, parameter) %>% summarize(
  means = mean(estimate)
)
M1parms <- (means %>% filter(Model == "M1"))$means
M2parms <- (means %>% filter(Model == "M2"))$means
M3parms <- (means %>% filter(Model == "M3"))$means
set.seed(4231352)
#M1sims1000 <- netvizinf::saom_simulate(dat = mydata,
#                            struct = M1eff,
#                       parms = as.numeric(M1parms),
#                       N = 1000)
#save(M1sims1000, file = "data/M1sims1000.RData")

# M2sims1000 <- netvizinf::saom_simulate(dat = mydata,
#                             struct = M2eff,
#                        parms = as.numeric(M2parms),
#                        N = 1000)
# save(M2sims1000, file = "data/M2sims1000.RData")

# M3sims1000 <- netvizinf::saom_simulate(dat = mydata,
#                             struct = M3eff,
#                        parms = as.numeric(M3parms),
#                        N = 1000)
# save(M3sims1000, file = "data/M3sims1000.RData")

@@

diff1 <- am_t1 - fd2.w1
diff1[which(diff1==1)] <- 2
diff1[which(fd2.w1 == 1 & am_t1 == 1)]  <- 1
diff1 <- diff1 %>% dplyr::as_data_frame() 
names(diff1) <- paste0("V", 1:16)
diff1 %>% mutate(from =paste0("V", 1:16)) %>% gather(to, value, -from) %>%
  mutate(from = as.character(parse_number(from)), to = as.character(parse_number(to)), 
         var = recode(value, `0` = "Neither", `2` = "InAvgNotW1",  `1`= "InBoth"),
         color2 = recode(value, `0`="white", `1`= "black", `2` = "green")) %>%
  filter(value !=0) %>% 
  full_join(t1, c("from","to")) -> t1_1 
t1_1 <- arrange(t1_1, from, to)
p1 <- ggplot(data = t1_1) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = t1_1$color2,
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
           singletons = T) +
  labs(title = "Difference between\nWave 1 and Average from M1") + ThemeNet +
  theme(plot.background = element_rect(color = 'black'))

diff2 <- -am_t1 + fd2.w2 
diff2[which(diff2==1)] <- 2
diff2[which(fd2.w2 == 1 & am_t1 == 1)]  <- 1
diff2 <- dplyr::as_data_frame(diff2)
names(diff2) <- paste0("V", 1:16)
diff2 %>% mutate(from =paste0("V", 1:16)) %>% gather(to, value, -from) %>% 
  mutate(from = as.character(parse_number(from)), to = as.character(parse_number(to)), 
         var = recode(value, `0` = "Neither", `-1` = "InAvgNotW2",  `1`= "InBoth", `2` = "InW2notAvg"),
         color2 = recode(value, `0`="white", `1`= "black", `2` = "red", `-1` = "green")) %>% 
  filter(value !=0) %>% 
  full_join(t1, c("from","to")) -> t1_2 
t1_2$linewidth[is.na(t1_2$linewidth) & !is.na(t1_2$to)] <- 1
t1_2 <- arrange(t1_2, from, to)
t1_2 <- t1_2 %>% filter(!(from == 16 & is.na(to)))
p2 <- ggplot(data = t1_2) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = t1_2$color2,
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
           singletons = T) +
  labs(title = "Difference between\nWave 2 and Average from M1") + ThemeNet +
  theme(plot.background = element_rect(color = 'black'))
grid.arrange(p1, p2, nrow = 1)
@

@@

% No longer using this plot
<<gof2, fig.cap="An example of viewing the distribution of a statistic of interest from a SAOM. Here, the Mahalanobis distance of the cumulative outdegree values computed on a simulated network to the mean value simulated is shown. The distance computed from the data is shown as a vertical line.", fig.height=3, out.width='.6\\textwidth', eval = FALSE>>=

library(lattice)
gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
histdat <- data.frame(x = log(gof1$`Period 1`$SimulatedTestStat))
ggplot(histdat) + geom_histogram(aes(x = x), binwidth = .1, fill = 'white', color = "black") +
  #geom_density(aes(x = x, y = ..density..*250)) +
  geom_vline(xintercept = log(gof1$`Period 1`$ObservedTestStat), color = 'red') +
  ThemeNoNet +
  labs(x = "log(Statistic Value)", title = "Distribution of log Mahalanobis Distance")

@@

\section{Collections of models}\label{sec:collectmodels}
